Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-01 young_pp_discs.tex
renamed

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from Ossenkopf and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies, giving a beam size close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
 \centering
 \begin{minipage}{80mm}
   \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
   \label{sigmas}
   \begin{tabular}{c||c|c|c|c}
     \hline
     Species & Transition & Frequency & Upper energy level & Sensitivity\\
     \hline
     C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
     OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
     HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
     H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ & 34.9$\,$K & 290.62GHz & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
     \hline
   \end{tabular}
 \end{minipage}
\end{table*}
In order to clean the images, which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in table \ref{sigmas}.\newline

@@

 \caption{CASA simulation on the continuum emission of the model at 300GHz, using the same ALMA configuration as used for the molecular lines. The size of the beam at 300GHz is shown in the lower left corner.}
 \label{continuum}
\end{figure}

>>>>

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies, giving a beam size close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
 \centering
 \begin{minipage}{80mm}
   \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
   \label{sigmas}
   \begin{tabular}{c||c|c|c|c}
     \hline
     Species & Transition & Frequency & Upper energy level & Sensitivity\\
     \hline
     C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
     OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
     HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
     H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62GHz & 34.9$\,$K & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
     \hline
   \end{tabular}
 \end{minipage}
\end{table*}
In order to clean the images, which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in table \ref{sigmas}.\newline

@@

 \caption{CASA simulation of the continuum emission of the model at 300GHz, using the same ALMA configuration as used for the molecular lines. The size of the beam at 300GHz is shown in the lower left corner.}
 \label{continuum}
\end{figure}
